The Evolution of Phenotypically Invariant Genetic Networks

### 2017-10-21 kalman_walk_paper.tex
added argument about evolution of seg variance

After selection, we have reproduction:
suppose this occurs as in the infinitesimal model \citep{infinitesimal},
so that each offspring of parents with traits $x$ and $y$
is drawn independently from a Gaussian distribution 
with mean $(x+y)/2$ and covariance matrix $N$.
Here, $N$ is the contribution of ``segregation variance''.
If $\widetilde \Sigma = (\Sigma^{-1} + U^{-1})^{-1}$ 
is the covariance matrix of the parents post-selection,
then the distribution of offspring will again be Gaussian,
with mean equal to that of the parents
and covariance matrix $\widetilde \Sigma/2 + N$.

In summary, a generation under this model modifies the mean ($\mu$)
and covariance matrix ($\Sigma$) of a population as follows:
\begin{align*}
    \mu &\mapsto \mu' = (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu \\
    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + N .
\end{align*}
What measures are stable under this transformation?
The condition $\mu = \mu'$ reduces to $\Sigma U^{-1} \mu = 0$;
if we assume $N$ and therefore $\Sigma$ are of full rank,
then this happens if and only if $\mu$ is not in the span of the columns of $U$,
i.e., if $\mu$ lies in a neutral direction.
The condition $\Sigma' = \Sigma$ can also be solved,
at least numerically.
After rearrangement, it reduces to
$\Sigma U^{-1} \Sigma + (I/2 - N U^{-1}) \Sigma = N$.
We can find a more explicit description
if we assume that $x^T U x = \sum_{i=1}^k x_i^2$,
i.e., that selection only cares about the first $k$ coordinates,
and then with no interactions between traits.
If so, the condition $\Sigma' = \Sigma$
can be written in block form as
\begin{align*}
    \begin{bmatrix}
        \Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11}
        & \Sigma_{11} \Sigma_{12} + (I/2 - N_{11})\Sigma_{12} \\
        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - N_{12}^T \Sigma_{11}
        & \Sigma_{22} - N_{12}^T \Sigma_{12} 
    \end{bmatrix}
    &=
    \begin{bmatrix}
        N_{11} & N_{12} \\
        N_{12}^T & N_{22}
    \end{bmatrix} .
\end{align*}
The first equation, $\Sigma_{11}^2 + (I/2 - N_{11}) \Sigma_{11} = N_{11}$,
can be solved with the quadratic formula: 
$$\Sigma_{11} = (N_{11} - I/2 + Q)/2$$
for any $Q$ that commutes with $N_{11}$
and is a solution to $Q^2 = (N_{11} - I/2)^2 + 4 N_{11}$.
Since we need $\Sigma$ to be positive definite,
we take the solution with positive eigenvalues.
Given $\Sigma_{11}$, the remaining components are
\begin{align*}
    \Sigma_{12} 
    &= 
    \left( \Sigma_{11} + I/2 - N_{11} \right)^{-1} N_{12} \\
    \Sigma_{22}
    &=
    N_{12}^T \Sigma_{12} + N_{22} .
\end{align*}
Importantly, the mean $\mu$ does not affect 
either how the covariance matrix moves,
or its stable shape.

Above we have described the \emph{expected} motion of the mean and covariance..
However, random resampling will introduce noise.
Suppose that a population of $N$ individuals
behaves approximately as described above.
By the above,
we may expect that the covariance matrix stays close to a constant value $\Sigma$,
computed from $N$ and $U$ as above,
so that we need only consider motion of the mean, $\mu$.
Since we take a sample of size $N$ to construct the next generation,
the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
and covariance matrix $\Sigma/N$.
Defining $\Gamma = (I - (I + \Sigma U^{-1})^{-1})$,
this can be written as
\begin{align*}
    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
\end{align*}
where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
and suppose that $\mu$ differs from optimal by something of order $1/N$:
if $\nu(t) = N \mu(tN)$,
then the previous equation implies that as $N \to \infty$, 
in the limit $\nu$ solves the It\^{o} equation
\begin{align*}
    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
\end{align*}
where now $W(t)$ is a multivariate white noise.
This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
\begin{align*}
    \nu(t) = e^{-t \Gamma} \nu_0 + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
\end{align*}
% Note that after some algebra, $\Gamma = \Sigma U^{-1} (I - \Sigma U^{-1})^{-1}$.
The asymptotic variance of this process in the direction $z$ is
\begin{align} \label{eqn:limiting_cov}
    \lim_{t \to \infty} \var[\nu(t) \cdot z]
    &=
    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
\end{align}
which is infinite iff $\Gamma z = 0$,
which occurs iff $Uz=0$.
In other words, population mean trait values lie away from the optimal set
by a Gaussian displacement of order $1/N$
with a covariance matrix given by equation \eqref{eqn:limiting_cov}.

@@

\section{Away from the optimum}
\label{apx:away_from_opt}

>>>>

After selection, we have reproduction:
suppose this occurs as in the infinitesimal model \citep{infinitesimal},
so that each offspring of parents with traits $x$ and $y$
is drawn independently from a Gaussian distribution 
with mean $(x+y)/2$ and covariance matrix $R$.
Here, $R$ is the contribution of ``segregation variance''.
If $\widetilde \Sigma = (\Sigma^{-1} + U^{-1})^{-1}$ 
is the covariance matrix of the parents post-selection,
then the distribution of offspring will again be Gaussian,
with mean equal to that of the parents
and covariance matrix $\widetilde \Sigma/2 + R$.

In summary, a generation under this model modifies the mean ($\mu$)
and covariance matrix ($\Sigma$) of a population as follows:
\begin{align*}
    \mu &\mapsto \mu' = (\Sigma^{-1}+U^{-1})^{-1} \Sigma^{-1} \mu \\
    \Sigma &\mapsto \Sigma' = \frac{1}{2} (\Sigma^{-1} + U^{-1})^{-1} + R .
\end{align*}
What measures are stable under this transformation?
The condition $\mu = \mu'$ reduces to $\Sigma U^{-1} \mu = 0$;
if we assume $R$ and therefore $\Sigma$ are of full rank,
then this happens if and only if $\mu$ is not in the span of the columns of $U$,
i.e., if $\mu$ lies in a neutral direction.
The condition $\Sigma' = \Sigma$ can also be solved,
at least numerically.
After rearrangement, it reduces to
$\Sigma U^{-1} \Sigma + (I/2 - R U^{-1}) \Sigma = R$.
We can find a more explicit description
if we assume that $x^T U x = \sum_{i=1}^k x_i^2$,
i.e., that selection only cares about the first $k$ coordinates,
and then with no interactions between traits.
If so, the condition $\Sigma' = \Sigma$
can be written in block form as
\begin{align*}
    \begin{bmatrix}
        \Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11}
        & \Sigma_{11} \Sigma_{12} + (I/2 - R_{11})\Sigma_{12} \\
        \Sigma_{12}^T \Sigma_{11} + \Sigma_{12}^T/2 - R_{12}^T \Sigma_{11}
        & \Sigma_{22} - R_{12}^T \Sigma_{12} 
    \end{bmatrix}
    &=
    \begin{bmatrix}
        R_{11} & R_{12} \\
        R_{12}^T & R_{22}
    \end{bmatrix} .
\end{align*}
The first equation, $\Sigma_{11}^2 + (I/2 - R_{11}) \Sigma_{11} = R_{11}$,
can be solved with the quadratic formula: 
$$\Sigma_{11} = (R_{11} - I/2 + Q)/2$$
for any $Q$ that commutes with $R_{11}$
and is a solution to $Q^2 = (R_{11} - I/2)^2 + 4 R_{11}$.
Since we need $\Sigma$ to be positive definite,
we take the solution with positive eigenvalues.
Given $\Sigma_{11}$, the remaining components are
\begin{align*}
    \Sigma_{12} 
    &= 
    \left( \Sigma_{11} + I/2 - R_{11} \right)^{-1} R_{12} \\
    \Sigma_{22}
    &=
    R_{12}^T \Sigma_{12} + R_{22} .
\end{align*}
Importantly, the mean $\mu$ does not affect 
either how the covariance matrix moves,
or its stable shape.

Above we have described the \emph{expected} motion of the mean and covariance..
However, random resampling will introduce noise.
Suppose that a population of $N$ individuals
behaves approximately as described above.
By the above,
we may expect that the covariance matrix stays close to a constant value $\Sigma$,
computed from $R$ and $U$ as above,
so that we need only consider motion of the mean, $\mu$.
Since we take a sample of size $N$ to construct the next generation,
the next generation's mean is drawn from a Gaussian distribution with mean $\mu'$
and covariance matrix $\Sigma/N$.
Defining $\Gamma = (I - (I + \Sigma U^{-1})^{-1})$,
this can be written as
\begin{align*}
    \mu' - \mu = \Gamma \mu + \epsilon/\sqrt{N} ,
\end{align*}
where $\epsilon$ is a multivariate Gaussian with mean zero and covariance matrix $\Sigma$.
Let $\mu(k)$ denote the mean in the $k^\text{th}$ generation,
and suppose that $\mu$ differs from optimal by something of order $1/N$:
if $\nu(t) = N \mu(tN)$,
then the previous equation implies that as $N \to \infty$, 
in the limit $\nu$ solves the It\^{o} equation
\begin{align*}
    d \nu(t) = \Gamma \nu(t) dt + \Sigma^{1/2} dW(t) ,
\end{align*}
where now $W(t)$ is a multivariate white noise.
This has an explicit solution as a multivariate Ornstein-Uhlenbeck process:
\begin{align*}
    \nu(t) = e^{-t \Gamma} \nu_0 + \int_0^t e^{-(t-s) \Gamma} \Sigma^{1/2} dW(s).
\end{align*}
% Note that after some algebra, $\Gamma = \Sigma U^{-1} (I - \Sigma U^{-1})^{-1}$.
The asymptotic variance of this process in the direction $z$ is
\begin{align} \label{eqn:limiting_cov}
    \lim_{t \to \infty} \var[\nu(t) \cdot z]
    &=
    \int_0^\infty z^T e^{-s \Gamma} \Sigma e^{-s \Gamma} z ds ,
\end{align}
which is infinite iff $\Gamma z = 0$,
which occurs iff $Uz=0$.
In other words, population mean trait values lie away from the optimal set
by a Gaussian displacement of order $1/N$
with a covariance matrix given by equation \eqref{eqn:limiting_cov}.

@@

%%%%%%%%%%%
\subsection*{Evolution of segregation covariance}

The description above does not completely describe how two diverging populations interact,
since the amount of \emph{segregation variance}, quantified by $R$,
will not stay constant.
To get an idea of how this might change,
suppose that a multivariate trait is determined by $L$ unlinked, biallelic loci,
and that the $i^\mathrm{th}$ locus has two alleles with additive effects $\pm x_i$,
so that begin homozygous for the $+$ allele contributes $+2x_i$ to the trait.
For simplicity, we will neglect the effects of selection.
\plri{fixup the below to be actually multivariate}
If the $+$ allele at locus $i$ is at frequency $p_i$ in a population,
then the mean and genetic variance of the trait in a diploid population with random mating is
\begin{align*}
    m &= 2 \sum_i p_i x_i \\
    s^2 &= 2 \sum_i p_i (1-p_i) x_i^2 .
\end{align*}

Segregation variance between two parents depends on the loci at which either are heterozygous,
and each locus contributes independently
since alleles are additive.
If the alleles are at Hardy-Weinberg proportions,
then since segregation is a fair coin flip, a heterozygous locus contributes $x_i^2/4$ to the variance,
and so the \emph{mean} segregation variance, averaging across parents, is
% two parents, prob 2p(1-p) for prob the locus is heterozygous, variance 0.5(1-0.5)x^2.
\begin{align*}
    R_0(p) = \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 .
\end{align*}

On the other hand,
if the second parent came from a distinct population
with frequencies $q_i$
(an $F_1$ hybrid),
this would be
\begin{align*}
    R_1(p,q) &= \frac{1}{4} \sum_i p_i^2 (1-p_i)^2 x_i^2 
                + \frac{1}{4} \sum_i q_i^2 (1-q_i)^2 x_i^2 \\
                &= (R_0(p) + R_0(q))/2 .
\end{align*}
If we assume that the populations are at equilibrium, $R_0(p) \approx R_0(q)$,
and so $R_1(p,q) \approx R_0(p)$.

Now consider an $F_2$ hybrid, where both parents are $F_1$
and so each heterozygous at locus $i$ with probability $p_i (1-q_i) + (1-p_i) q_i$.
Then
\begin{align*}
    R_2(p,q) &= \frac{1}{4} \sum_i \left(p_i (1-q_i) + (1-p_i) q_i \right) x_i^2  .
\end{align*}
Suppose that the two populations are slightly drifted from each other,
with frequency difference $p_i-q_i = 2\epsilon_i$.
Then,
$$
    p (1-q) + p (1-q)
    &=
    (u+\epsilon) (1-u+\epsilon) + (u-\epsilon) (1-u-\epsilon) \\
    &=
    2 u (1-u)
    % + \epsilon (u + (1-u) - u - (1-u))
    + 2 \epsilon^2 .
$$
If the frequencies have evolved neutrally in unconnected, Wright-Fisher populations 
of effective size $N$ for $t$ generations from a common ancestor with allele frequency $u$,
then $\epsilon$ has mean zero and variance roughly $u (1-u) t/N$.
Still assuming the populations are at stationarity,
so that $R_0$ is constant between the two,
and taking the frequencies $p_i$ as a proxy for the ancestral frequencies $u_i$,
this implies that we expect
\begin{align*}
    R_2 
    &\approx 
    R_0 + \frac{1}{2} \sum_i p_i (1-p_i) x_i^2 t/N \\
    &=
    \left( 1 + \frac{t}{N} \right) R_0 .
\end{align*}
On the other hand, the expected squared difference in trait \emph{means} here is
\begin{align} \label{eqn:var_dx}
    4 \sum_i p_i (1-p_i) x_i^2 t / N = 8 R_0 t /N .
\end{align}
This implies that under this model,
segregation variance in $F_2$s between two populations
is roughly increased by a factor of $1/8$ of the difference between their means.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Away from the optimum}
\label{apx:away_from_opt}
